Customizing Software by Direct Manipulation of Tabular Data

### 2020-05-22 paper.tex
finish section 3

\emph{Making edits}: The query engine can issue a request to a table
adapter to make an edit to a record. The meaning of making an edit can
vary depending on the adapter: in the local storage adapter, an
annotation may be persisted into local storage; in the DOM scraping
adapter, an edit may represent filling in a form field.

@@

After specifying basic metadata like which URLs the adapter should apply
to, the programmer only needs to implement a single Javascript function
that scrapes relevant elements from the page. They are free to use any
scraping techniques; in practice simple CSS selectors and DOM APIs
usually suffice. he generic adapter then wraps this scraping function to
implement the table adapter interface. For example, when the table view
is sorted, the generic adapter takes the DOM elements corresponding to
the rows in the table, removes all of them from the page, and then
reinserts them in the new order.

@@

In certain types of software, customizability is often touted as a key
selling point. An integrated website adapter would provide a way for
developers to integrate with an existing ecosystem of formulas and
customization tools, without needing to build all that functionality
from scratch.

\textbf{Shared storage adapter}: It would be useful to share user
annotations among users and across devices---for example,
collaboratively taking notes with friends on options for places to stay
on Airbnb. The existing Local Storage Adapter could be extended to share
live synchronized data with other users. This could be achieved through
a centralized web server, or through P2P connections that might provide
stronger privacy guarantees.

@@

Queries are processed in three steps. First, the query invokes a primary
DOM scraping table adapter that associates records in the result with
elements in the application's user interface. At minimum, the primary
table adapter needs to return record IDs and have the ability to
manipulate the application's UI. It can also optionally return data
about each record. Next, additional tables (AJAX data, local storage
data) are left-joined by ID. Finally, the result table can be sorted and
filtered by any column.

This query model can be viewed as a tiny (and rather constrained) subset
of the SQL query model. Despite its simplicity, this simple model has
proven sufficient for meeting the needs of customization in practice,
and minimizes the complexity of supporting more general and arbitrary
queries. But because it fits into the general paradigm of relational
queries, it could theoretically be extended to support a wider range of
queries.

The query engine is also responsible for executing formulas. We have
built a small formula language resembling the ones used in visual query
tools like SIEUFERD. As in those tools, and unlike in spreadsheets,
formulas automatically apply across an entire column of data, and
reference other column names instead of values on specific rows. For
example, the user would write \texttt{=price\ *\ 2} to create a column
with a value derived from a price attribute. This is more convenient
than needing to copy-paste a formula across an entire column as in
spreadsheets, and none of our customization use cases have uncovered a
need for writing different formulas on different rows of the table.

@@

In Wildcard we provide a basic table editor as the user interface on top
of the query engine. It is built with the Handsontable Javascript
library, which provides UI elements for viewing and editing a table, as
well as basic query operations like sorting and filtering.

\emph{todo: other table instruments}

@@

(\emph{note: written with the assumption that Expedia gets cut from the
Examples section})

Cell editors are UI widgets that expose a custom editing UI for a single
cell of the table view. A programmer building a cell editor need only
integrate it with the table viewer; propagating values into the website
UI is handled by the site-specific DOM adapter. As a result, creating a
new cell editor can be as easy as just writing some glue code between
the table editor and an existing UI widget library.

>>>>

\emph{Handling edits}: The query engine can issue a request to a table
adapter to make an edit to a record. The meaning of making an edit can
vary depending on the adapter: in the local storage adapter, an
annotation may be persisted into local storage; in the DOM scraping
adapter, an edit may represent filling in a form field.

@@

After specifying basic metadata like which URLs the adapter should apply
to, the programmer only needs to implement a single Javascript function
that scrapes relevant elements from the page. They are free to use any
scraping techniques; in practice simple CSS selectors and DOM APIs
usually suffice. The generic adapter then wraps this scraping function
to implement the table adapter interface. For example, when the table
view is sorted, the generic adapter takes the DOM elements corresponding
to the rows in the table, removes all of them from the page, and then
reinserts them in the new order.

@@

Customizability is sometimes a key selling point for software. An
integrated website adapter would provide a way for developers to
integrate with an ecosystem of formulas and customization tools without
needing to build all that functionality from scratch.

\textbf{Shared storage adapter}: It would be useful to share user
annotations among users and across devices---for example,
collaboratively taking notes with friends on options for places to stay
together on a trip. The existing local storage adapter could be extended
to share live synchronized data with other users. This could be achieved
through a centralized web server, or through P2P connections that might
provide stronger privacy guarantees.

@@

Queries are processed in three steps. First, the query invokes a primary
DOM scraping adapter that associates table rows with elements in the
application's user interface. Next, additional tables (AJAX data, local
storage data) are left-joined by ID. Finally, the result table is sorted
and filtered according to user-specified predicates.

One way to view this query model is as a tiny subset of the SQL query
model. Despite its simplicity, this model has proven sufficient for
meeting the needs of customization in practice, and minimizes the
complexity of supporting arbitrary queries. But because it fits into the
general paradigm of relational queries, it could theoretically be
extended to support a wider range of queries.

The query engine is also responsible for executing formulas. We have
built a small formula language resembling a spreadsheet formula
language, except that formulas automatically apply across an entire
column of data, and reference other column names instead of values in
specific rows. This is more convenient than needing to copy-paste a
formula across an entire column as in spreadsheets, and has worked for
all of the customizations we have built.

@@

We provide a basic table editor as the user interface on top of the
query engine. Our table editor is built with the Handsontable Javascript
library, which provides UI elements for viewing, editing, sorting, and
filtering a table.

In addition to the basic table editing operations, we also provide
\emph{cell editors}: UI widgets that expose a custom editing UI for a
single cell of the table view. A programmer building a cell editor need
only integrate it with the table viewer; propagating values into the
website UI is handled by the site-specific DOM adapter. In
Section~\ref{sec:evaluation} we provide some examples of using cell
editors.

The table editor only serves as a shallow interface layer over the query
engine, relaying user commands to the query engine and rendering the
resulting data table. Because of this architectural split, it would be
straightforward to develop additional table editor interfaces on top of
the Wildcard system. One intruiging future possibility is to apply the
idea of cell editors to the entire table, by developing table editors
which don't look like a table. For example, we could provide a calendar
view for displaying a table containing a date column.

@@
